---
title: "RNASeq Analysis"
author: "Otoniel Maya"
date: "16-Feb-2024"
format: html
editor: visual
execute: 
  eval: false
---

# Reporting data analysis

For this analysis try to keep a estructured folder tree in the same directory. Something like this:

```{r, eval=FALSE}
project_folders <- function() {
  base::dir.create("00_raw_data")
  base::dir.create("01_analysis")
  base::dir.create("02_tidy_data")
  base::dir.create("03_r_scripts")
  base::dir.create("04_plots")
  base::dir.create("99_results")
}

# project_folders()
```

You can follow your own system but keep a logical and sequential order. For this tutorial we will follow this convention using Snakemake inside the `analysis` folder.

This tutorial is a reproducible implementation from: [Research Technology Bioinformatics Tools for Life Science RNAseq tutorial](https://tuftsdatalab.github.io/Research_Technology_Bioinformatics/tutorials/IntroToRNAseq/lessons/05_Differential_Expression.html) to make it reproducible™.

### Packages

Initiate `renv`:

```{r, echo=FALSE}
#| warning: true

if (!require("renv", quietly = TRUE))
    install.packages("renv", ask=FALSE)

renv::init(bioconductor = TRUE, force = TRUE)
```

### CRAN Packages installation on renv

```{r}
#| warning: false


pkgs <- c("here", "tidyverse", "yaml")

# Install the missing packages, including their dependencies
for (lib in pkgs[!pkgs %in% installed.packages()]) {
    install.packages(lib, dependencies = TRUE)
}
```

Bioconductor missing dependency to handle Kallisto h5 files.

```{r, echo=FALSE}
#| warning: false

BiocManager::install('rhdf5', update = FALSE)
# BiocManager::install(c('SummarizedExperiment', 'HDO.db', 'org.Sc.sgd.db', 'DESeq2', 'apeglm', 'tximport', 'rhdf5', 'biomaRt'), update = FALSE)

# renv::status()
```

Usually `renv::init()` creates a renv.lock file but if something failes it will be necessary to take a renv snapshot manually:

```{r}
renv::snapshot()
```

Loading all the packages, however, to keep track of each function a syntax `library::function()` will be used in the follow code.

```{r}
#| warning: false


library(DESeq2)
library(ggplot2)
library(dplyr)
library(magrittr)
library(DEGreport)
library(pheatmap)
library(org.Sc.sgd.db)
library(clusterProfiler)
```

### Loading data

Don't forget to use here, it helps a lot with **`here::here()`**

```{r}
metadata <- readr::read_table(here::here("analysis/metadata.tsv"))

kallisto_dir <-"analysis/rnaseqSE_snakemake/02_mapping/kallisto"

lsdir <- list.dirs(here::here(kallisto_dir), recursive = FALSE)

kallisto_files <- paste0(lsdir, "/abundance.h5")
```

```{r}
ensembl <- biomaRt::useMart(biomart = "ensembl", 
                            dataset = "scerevisiae_gene_ensembl")

t2g <- biomaRt::getBM(attributes = c("ensembl_transcript_id",
                                     "ensembl_gene_id",
                                     "external_gene_name"), 
                      mart = ensembl)

txi_kallisto <- tximport::tximport(kallisto_files, 
                                   type = "kallisto", 
                                   tx2gene = t2g, 
                                   ignoreTxVersion = TRUE) 
```

```{r}
dds <- DESeq2::DESeqDataSetFromTximport(txi_kallisto, colData=metadata, design = ~ condition)

dds <- DESeq2::DESeq(dds)
```

### PCA analysis

```{r}
rld <- DESeq2::rlog(dds, blind=TRUE)
DESeq2::plotPCA(rld, intgroup="condition") + ggplot2::geom_text(ggplot2::aes(label=name))
```

## Creating contrasts and running a Wald test

```{r}
contrast <- c("condition", "SNF2", "WT")
res_unshrunken <- DESeq2::results(dds, contrast=contrast)
DESeq2::summary(res_unshrunken)
```

### Shrinkage of the log2 fold changes

```{r}
res <- DESeq2::lfcShrink(dds, contrast=contrast, res=res_unshrunken, type="normal")
DESeq2::summary(res)
head(res)
```

### Filtering to find significant genes

```{r}
padj_cutoff <- 0.05 # False Discovery Rate cutoff
significant_results <- res[which(res$padj < padj_cutoff), ]
```

save results using customized file_name

```{r}
write.table(significant_results, 
            'analysis/significant_padj_0.05.txt',
            quote=FALSE)
```

### Visualization

Simple plot for a single gene YOR290C (SNF2)

```{r}
plotCounts(dds, gene="YOR290C", intgroup="condition")
```

### Heatmap

Plot multiple genes in a heatmap: extract the counts from the rlog transformed object and select by row name using the list of genes:

```{r}
significant_results_sorted <-  significant_results[order(significant_results$padj), ]
significant_genes_25 <- rownames(significant_results_sorted[1:25, ])

rld_counts <-  SummarizedExperiment::assay(rld)
rld_counts_sig <- rld_counts[significant_genes_25, ]
colnames(rld_counts_sig) <- metadata$sample

annotation <- as.data.frame(metadata)
rownames(annotation) <- annotation$sample
annotation$sample <- NULL

pheatmap::pheatmap(rld_counts_sig,
         cluster_rows = T,
         show_rownames = F,
         annotation_col = annotation,
         border_color = NA,
         fontsize = 10,
         scale = "row",
         fontsize_row = 8,
         height = 20)
```

load previously saved result

```{r}
significant_results_test <- read.table("analysis/significant_padj_0.05.txt", 
                                  header=TRUE, row.names = 1)
```

### Volcano plot

Add another column in the results table to label the significant genes using threshold of padj\<0.05 and absolute value of log2foldchange \>=1

```{r}
#| echo: true

res_table <- res %>%
    data.frame() %>%
    tibble::rownames_to_column(var="gene") %>%
    tibble::as_tibble()

res_table <- res_table %>%
    dplyr::mutate(threshold_OE =  padj < 0.05 & abs(log2FoldChange) >= 1)

# you can view the modified table
head(res_table)
```

make volcano plot, the significant genes will be labeled in red

```{r}
ggplot(res_table) +
    geom_point(aes(x = log2FoldChange, 
                   y = -log10(padj), 
                   colour = threshold_OE)) +
    scale_color_manual(values=c("black", "red")) +  # black v.s. red dots
    ggtitle("SNF2 against WT") +                       # this line defines the title of the plot
  xlab("log2 fold change") +                      # this line defines the name of the x-axis
  ylab("-log10 adjusted p-value") +               # name of y-axis
  scale_x_continuous(limits = c(-7.5,7.5)) +      # the axis range is set to be from -7.5 to 7.5
  theme(legend.position = "none", #c(0.9, 0.9),
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))  
```

### Functional analysis

Run GO enrichment analysis for the top 500 genes using clusterProfiler

```{r}
significant_results_sorted <- res[order(res$padj), ]
significant_genes_500 <- rownames(significant_results_sorted[1:500, ])
ego <- clusterProfiler::enrichGO(gene = significant_genes_500,
                                 keyType = "ENSEMBL",
                                 OrgDb = org.Sc.sgd.db::org.Sc.sgd.db)
```

Output results from GO analysis to a table

```{r}
cluster_summary <- data.frame(ego)
```

Dotplot for the top 50 genes

```{r}
clusterProfiler::dotplot(ego, showCategory=10)
```

Enrichmap clusters the 50 most significant (by padj) GO terms to visualize relationships between terms

```{r, eval=FALSE}
d <- GOSemSim::godata("org.Sc.sgd.db", ont = "BP")    
compare_cluster_GO_emap <- enrichplot::pairwise_termsim(ego,
                                                        semData = d,
                                                        method="Wang")

clusterProfiler::emapplot(compare_cluster_GO_emap, showCategory = 10)
```
